---
title: "Introduction to R and interactive programming"
author: "Marc, Sean, Bronson and Debbie"
date: "10/7/2015"
output: html_document
---

Welcome to the second course in our series of courses designed to help train Biologists who want to be able to analyze their own data.  This course will introduce you to the experience of using high level interactive programming languages and specifically: to R.  In this class we will be introducing you to the language as a tool that you can use to explore data.  The course after this will attempt to teach you enough to actually get started with writing your own scripts etc.

# Introduction to R

### Getting familiar with RStudio
First lets get you used to RStudio.  RStudio is one of the most popular tools for working with R today.  It normally has four main panels: 'Console', 'Source', 'Environment', and 'Files'.  Lets take a minute and look at each of these:

#### Console panel:
This panel is your actual R console.  This panel is basically what you would see if you were to run R at the command line without any of the RStudio bells and whistles.

#### Source panel:
The Source panel is where you can edit scripts and markdown documents etc.  It's primary function is to be a text editor.  You can use any text editor that you like to write R, but this one is nice because it comes pre-configured with sensible defaults for source code highlighting.  This makes it easier to read your source code.  Also convenient: if you highlight a command and then hit CTRL-ENTER keys, it will send the highlighted text directly to the console to be executed.

#### Environment panel:
This panel is where you will see important details about your R session.  Any local variables will be shown here, and there is also a tab to see your command history.  This can be convenient if you are debugging or need to look at what is in a local object.

#### Files panel:
The Files panel has many tabs.  The Files tab will let you look at the files in the current working directory.  And the plots tab is where RStudio will draw plots that you generate. The Packages tab will let you browse installed packages to look at their helpful vignettes, and the help tab is where RStudio will render manual pages that you open.  The Viewer tab has a similar role to the plots tab, except that it's usually used for looking at interactive shiny widgets when using RStudio.



### Getting help 
If you forget everything else we teach you today, I want to make sure that you know how and where to look for help. The next set of topics will cover that because R has a large number of ways that new users can get assistance for learning about it.

#### Man pages
The manual page is usually the first place you will look if you are wondering what an R command will do.  R man pages will explain both what something is, and how to use it  You can usually pull up a manual page for a command by typing `?` followed by the command.  Like this:

```{r, eval=FALSE}
?data.frame
```

However a good habit is to wrap the thing you want help on in quotes like this:

```{r, eval=FALSE}
?'data.frame'
```

Why?  Because those quotes will allow you to search for help on strings that have special characters like dashes in them.  Why would you want to do that when this is not allowed for function names?  Consider the following:

```{r, eval=FALSE}
 ?'GenomicRanges-class'
```

The above will take you to the help page for the GenomicFeatures S4 class.  It's a very useful man page, but you can't get there if you leave out the quotes.

Alternatively, you could choose to just use the `help` command.  But that is three extra letters of typing (and you would still have to quote the arguement)...

##### example()
One of the more useful (and often overlooked) functions for exploring a new function is the `example` function.  The `example` function takes advantage of the fact that most man pages have an example section.  So if instead of calling `help`, you call `example`, the examples from the corresponding manual page will be run for you in your R session.

```{r}
example('data.frame')
```



#### vignettes
One step up from manual pages are vignettes.  There are two kinds of vignettes for any R package.  One is the automatically generated kind.  This kind consists of all the manual pages collated together under a table of contents.  This kind of vignette is not especially useful.  The second kind of vignette though is very useful.  This second kind of vignette gives an overview of how the various kinds of functions and objects were intended to be used.  Many of the best R packages will contain this second, more useful kind of vignette.  When they are available, these are a great way to see what the package authors intention was.

#### Workflows
Sometimes communities will organize around a series of packages that work well together (usually not an accident).  When this happens you can sometimes find another kind of document to explain how these different packages are best used together.  This is called a workflow, and it's a great place to start if you are looking to learn about a new set of tools.

#### Forums
All software has bugs.  But open sourced software often will allow you to write to a community of like minded users (which often includes the package authors) to get help.  There are many valuable forums where you can get questions answered about software.  Always search first though.  Most of the time you will find that someone else has already asked your question and gotten an answer.

#### Cheat sheets
One of the most challenging things for new users of R is not getting help on commands that they know about.  It's usually knowing what the command is called in the first place (so that you can start to look for some help about it).  A number of R 'cheat sheets' are available online that can help with this problem.  Here is one of the more popular ones:

```
https://cran.r-project.org/doc/contrib/Short-refcard.pdf
```
 
This cheat sheet is valuable enough for new students that I plan to give you a copy of this cheat sheet, but there are many, many others on the web that cover a range of topics...


### calling functions
Now that we have discussed how to get help lets talk about basic R functions. 

#### For R functions, arguments can either be passed in to them "in order"

```{r}
someFun(val1, val2, val3)
```

#### OR by name

```{r}
someFun(arg1=val1, arg2=val2, arg3=val3)
```


#### OR as a mixture (but once you start using names, you had better stick with it).

```{r}
someFun(arg1, arg2, arg3=val3)
```

#### Remember to always check the man page
Some functions may not behave in the way you expect.


### Getting data into R 
#### read.table
#### scan
#### Other options: Databases, URIs, REST, HDF5 and many more


### Atomic vectors and other fun Data structures
#### Vectors, length
#### Lists
#### data.frames, matrices, dim
#### S4 objects


### Data object management
#### class()
#### as.list() and friends
#### subsetting operators [] vs [[]]
#### data casting
#### match(), %in%, unique()

### Flow control
#### if, else
#### switch


### Exploring packages
#### CRAN 
#### Bioconductor
#### github


### Bioconductor
#### Walkthroughs, vignettes and man pages.






